A Fast and Tight Heuristic for A* in Road Networks

### 2021-10-12 paper/ch_potentials.tex
lots of rough writing

@@

Truck routing differs from car routing due to night driving bans and other restrictions.
In~\cite{kswz-erptd-p-20}, a preliminary version of CH-Potentials\footnote{
In~\cite{kswz-erptd-p-20}, CH-Potentials are used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
Our submitted paper is the finished version of the ArXiv preprint.
% der punkt macht fuer mich keinen Sinn.
% Dass wir keine der Contributions von kswz-erptd-p-20 beschreiben ist ja offensichtlich, sieht ja jeder in dem Abschnitt.
% Wichtiger finde ich, dass kswz-erptd-p-20 nix von uns beschreibt.
% Das meinte ich zumindest mit dem Satz den du geloescht hast.
% Aber der jetzt folgende Satz sagt nur, dass unser paper hier nix von kswz-erptd-p-20 beschreibt.
\cite{kswz-erptd-p-20} does not describe any of the contributions of this paper.
} is used for such a scenario.
The work considers time-dependent blocked edges and waiting at parking locations.
Further, a trade-off between arrival time and route quality is considered.

@@

\begin{table}
\centering
\caption{Instances used in the evaluation.}\label{tab:graphs}
\input{table/graphs.tex}
\end{table}

@@

\paragraph{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
The graph has 16M nodes and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://mapbox.com}}.
The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
The PTV instances are not OSM-based.
One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
Table~\ref{tab:graphs} contains an overview over our instances.
In this table, we further include the sequential running time necessary to construct the CH.
We report preprocessing running times as averages over 10 runs.
For queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.

\paragraph{Experiments}

@@

@@

The performance of A* depends on the tightness of the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
For $\alpha > 1.1$, the growth slows down.
This illustrates both the strengths and limits of our approach and goal directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for a sufficiently tight lower bounds at preprocessing time.

@@

Table~\ref{tab:building_blocks} depicts the performance of A* with different heuristics and optimizations.
We compare CH-Potentials to three other heuristics.
First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
This corresponds to using Dijkstra's algorithm.
Second, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Our ALT implementation is uni-directional.
In this work, we do not consider bidirectional search as it creates problems for some settings, such as predicted traffic.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

We observe that the number of queue pushes roughly correlates with running time.
Each optimization reduces both queue pushes and running times.
All optimizations yield a combined speed-up of around 3.
% Skipping degree three nodes has the smallest impact.
CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly fewer nodes.
This is not surprising, since ALT computes worse distance estimates.
In contrast, CH-Potentials already compute exact distances with respect to $w_\ell$.
The number of popped nodes is the same for CH-Potentials and Oracle-A*.
The only difference between CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
This overhead leads to a slowdown of around 1.6.
Thus, CH-Potentials are already very close to the best possible heuristic in this model.
% Very interesting is the comparison with Oracle-A*.
% Our algorithm is only 1.6 times slower than accessing the exact $w_\ell$ distances instantly from a precomputed array.
This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.

\begin{table}
\centering
\caption{
TODO
OSM Ger, Weight Scaled uniformly by 1.05, alternating dirs
}\label{tab:bidir_pruning}
\input{table/bidir_pruning.tex}
\end{table}

\begin{table}
\centering
\caption{
TODO
OSM Ger, Weight Scaled 1.5, sym with improved pruning, avg without, with topo opt
}\label{tab:bidir_switching}
\input{table/bidir_switching.tex}
\end{table}

\begin{table}
\centering
\caption{
TODO
OSM Ger
}\label{tab:bidir}
\input{table/bidir.tex}
\end{table}

@@

\bibliography{references}
\end{document}

>>>>

% TODO alternating directions

@@

Truck routing differs from car routing due to night driving bans and other restrictions.
In~\cite{kswz-erptd-p-20}, a preliminary version of CH-Potentials\footnote{
In~\cite{kswz-erptd-p-20}, CH-Potentials are used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
Our submitted paper is the finished version of the ArXiv preprint.
\cite{kswz-erptd-p-20} does not describe any of the contributions of this paper.
} is used for such a scenario.
The work considers time-dependent blocked edges and waiting at parking locations.
Further, a trade-off between arrival time and route quality is considered.

@@

\begin{table}
\centering
\caption{Instances used in the evaluation with preprocessing running times to construct a CH or CCH-Potential. With CCH-Potentials $w_{\ell}$ can be updated by rerunning Phase 2 again.}\label{tab:graphs}
\input{table/graphs.tex}
\end{table}

@@

\subsection{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
The graph has 16M nodes and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://mapbox.com}}.
The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACs implementation challenge~\cite{DemetrescuGJ09}.
Additionally, we have three graphs with proprietary traffic predictions also provided by PTV.
The PTV instances are not OSM-based.
One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
% TODO PTV20
Table~\ref{tab:graphs} contains an overview over our instances.
In this table, we further include the sequential preprocessing running time to construct the (C)CH.
We report preprocessing running times as averages over 10 runs.
For queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.

% Methodology
% time per source
% selection times

% what do we want to show
% - for one source target set we are as fast as RPHAST
%   - dimensions: phase (style), graph (color), CH Var (row)
% - incremental is cool

@@

\subsubsection{k-Nearest-Neighbor Queries with CCH}

One-to-many queries appear in many extended applications as a subproblem.
In some of these cases Lazy RPHAST can be used to achieve significant speed-ups with surprisingly little effort.
One such example are k-Nearest-Neighbor Queries with CCH~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
There, shortest distances are computed repeatedly from the same source to boundary nodes of a cell of nodes.
On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code~\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
We reran their improved experiments and include the results in our appendix.
Briefly summarized, integrating Lazy RPHAST into this algorithm decreased running times by up to 15 times depending on the distribution of the target set for almost no effort.

\subsection{(C)CH-Potentials Heuristic}

@@

The performance of A* depends on the tightness of the heuristic and the overhead of evaluating the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
For $\alpha > 1.1$, the growth slows down.
This illustrates both the strengths and limits of our approach and goal directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for a sufficiently tight lower bounds at preprocessing time.

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
We compare (C)CH-Potentials to three other heuristics.
First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
This corresponds to using Dijkstra's algorithm.
Second, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

We observe that the number of queue pushes roughly correlates with running time.
Each optimization reduces both queue pushes and running times.
All optimizations yield a combined speed-up of around 3.
% Skipping degree three nodes has the smallest impact.
CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly fewer nodes.
This is not surprising, since ALT computes worse distance estimates.
In contrast, CH-Potentials already compute exact distances with respect to $w_\ell$.
The number of popped nodes is the same for CH-Potentials, CCH-Potentials and Oracle-A*.
The only difference between (C)CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
This overhead leads to a slowdown of around 1.6.
Thus, CH-Potentials are already very close to the best possible heuristic in this model.
% Very interesting is the comparison with Oracle-A*.
% Our algorithm is only 1.6 times slower than accessing the exact $w_\ell$ distances instantly from a precomputed array.
This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
CCH-Potentials are slightly slower than CH-Potentials because their search space is denser.

\begin{table}
\centering
\caption{Performance of different variants of bidirectional A* on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:bidir_pruning}
\input{table/bidir_pruning.tex}
\end{table}

Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A*.
As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
Symmetric bidirectional A* without our improved pruning has the worst performance.
Enabling the improved pruning brings the performance symmetric bidirectional A* on par with the average potential for all heuristics except ALT.
Apparently, the average potential gives ALT a better heuristic than the original one for for each direction individually, contrary to the statements made in~\cite{gh-cspas-05}.
For all other heuristics, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
Without the low degree improvements, the improved symmetric variant is marginally faster and with the low degree improvements the average potential remains slightly faster.
This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.

\begin{table}
\centering
\caption{
Performance of bidirectional and unidirectional A* on OSM Ger with different query weights.
The symmetric variant uses the improved pruning, the average variant does not.
All variants use all low degree optimizations.
}\label{tab:bidir}
\input{table/bidir.tex}
\end{table}
% TODO remove fourth scenario

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search for A* depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
When the query weights are equal, CH-Potentials will already only traverse the shortest path with unidirectional search.
In this case it is clear that bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes significantly when the query weight increases are applied non-uniformly as in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
This makes bidirectional search so critical for ALTs performance.
We conclude that having a potential as tight as CH-Potentials usually makes bidirectional search unnecessary.
Bidirectional search for A* only pays of when the reduced weights exhibit certain very specific patterns.

\subsection{Applications}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix

\section{Improved Results for k-Nearest-Neighbor Queries on CCH}

\begin{table}
\centering
\caption{
Performance of different closest-POI algorithms for various POI distributions.
For each distribution, we report the time to index a set of POIs (selection time), the space consumed by the
index (selection space), and the time to find the $k = 1, 4, 8$ closest POIs (query time).
}\label{tab:knn}
\input{table/knn.tex}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig/knn_ball_size.pdf}
\caption{
Selection and query times of various closest-POI algorithms with $|P| = 2^14$ POIs picked at random from a ball of varying size $|B|$.
Queries find the $k = 4$ closest POIs.
}\label{fig:knn_ball_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig/knn_num_pois.pdf}
\caption{
Selection and query times of various closest-POI algorithms with a varying number $|P|$ of POIs picked at random from a ball of size $|B| = |V|$.
Queries find the $k = 4$ closest POIs.
}\label{fig:knn_num_pois}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.6363\linewidth]{fig/knn_demands.pdf}
\caption{
Time to generate a single trip with different demand generators for various values of $\lambda$.
}\label{fig:knn_demands}
\end{figure}

\section{Extended Experimental Results on Bidrectional A*}

\begin{table}
\centering
\caption{
Performance of different direction selection criteria of bidirectional A* on OSM Ger with different query weights.
The symmetric variant uses the improved pruning, the average variant does not.
All variants use all low degree optimizations.
}\label{tab:bidir_switching}
\input{table/bidir_switching.tex}
\end{table}

\end{document}
